---
title: "Bayside Property Analysis - AD 571 Assignment 1"
format:
  html: default
date: "2025-07-08"
execute:
  eval: false
---

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(scales)
BOROUGH <- read_csv("BOROUGH.csv")
BUILDING_CLASS <- read_csv("BUILDING_CLASS.csv")
NEIGHBORHOOD <-read_csv("NEIGHBORHOOD.csv")
NYC_TRANSACTION_DATA <- read_csv("NYC_TRANSACTION_DATA.csv")

#JOIN DATA using left join <- matching variable names
all_nyc <- NYC_TRANSACTION_DATA %>% left_join(BUILDING_CLASS, by =c("BUILDING_CLASS_FINAL_ROLL" =
                                                                     "BUILDING_CODE_ID")) %>% left_join(NEIGHBORHOOD, by=c("NEIGHBORHOOD_ID" = "NEIGHBORHOOD_ID"))
```

```{r}
3.
bayside <- filter(all_nyc, NEIGHBORHOOD_NAME == "BAYSIDE")
nrow(bayside)

bayside_residential <- bayside%>%
  filter(TYPE == "RESIDENTIAL")
nrow(bayside_residential)
```

```{r}
4.
nrow(bayside_residential)

ggplot(bayside_residential, aes(x = SALE_PRICE)) +
  geom_histogram(fill = "skyblue", color = "black", bins = 20, boundary = 0) +
  scale_x_continuous(labels = scales::dollar, limits = c(0, 2000000))+
  labs(
    title = "Distribution of Sale Prices of Properties in Bayside, NYC",
    x = "Sale Price ($)",
    y = "Number of Sales"
  ) +
  theme_minimal()
```
This graph shows a very high number (over 5000) of properties with a sale price of $0, with a uniform distribution to the sale price of  750,000, then it skews out to the right-hand side. The unimodal peak at $0 suggests that there are a large number of properties entered with no sale value or placeholder values. However, with these $0 sales, the graph does not accurately display the distribution of the sales. 
```{r}
bayside_residential_greater_than_0 <- bayside%>%
  filter(SALE_PRICE > 0)

nrow(bayside_residential_greater_than_0)

ggplot(bayside_residential_greater_than_0, aes(x = SALE_PRICE)) +
  geom_histogram(fill = "skyblue", color = "black", bins = 20) + 
  scale_x_continuous(labels = scales::dollar, limits = c(0,2000000))+
  labs(
    title = "Distribution of Sale Prices (Greater than 0) of Properties in Bayside, NYC",
    x = "Sale Price ($)",
    y = "Number of Sales"
  )+
theme_minimal()
```
With the sales of $0 now removed, we can get a more accurate representation of the data in the bayside area. The main cluster of the data is between $200,000 and $800,000, then skews out to the right hand side where there are a smaller number of properties with high-value sales. 

The main difference with the removal of property sales equaling $0 is that it provides a much more accurate representation of the property distribution. The $0 sales were skewing the datas representation on the graph, with a large drop off of properties with an actual sale value. The genuine sales gave a higher level of information on the distribution. 
```{r}
5.
bayside_residential_greater_than_0$SALE_DATE <- as.Date(bayside_residential_greater_than_0$SALE_DATE, format = "%Y-%m-%d")

end_date <- max(bayside_residential_greater_than_0$SALE_DATE, na.rm = TRUE)
start_date <- end_date - 365

bayside_latest_year <- bayside_residential_greater_than_0%>%
  filter(bayside_residential_greater_than_0$SALE_DATE >= start_date & SALE_DATE <= end_date)

total_sales <- sum(bayside_latest_year$SALE_PRICE, na.rm = TRUE)
company_revenue <- total_sales * 0.125 * 0.05

print(paste("Total Sales and Revenue from", start_date,"to", end_date))
print(paste("Total Sales = ",dollar(total_sales)))
print(paste("Expected Company Revnue = ",dollar(company_revenue)))
```
```{r}
6.
bayside_latest_year$QUARTER <- quarter(bayside_latest_year$SALE_DATE)
bayside_latest_year$SEASON <- factor(
  bayside_latest_year$QUARTER,
  levels = 1:4,
  labels = c("Winter", "Spring", "Summer", "Fall")
)

ggplot(bayside_latest_year,aes(x = SEASON, fill = SEASON)) +
  geom_bar() +
  labs(
    title = "Residential Property Sales in Bayside by Season, Lastest Year",
    subtitle = paste(
      format(start_date, "%b %d, %Y"), "to", format(end_date, "%b %d, %Y")
      ),                                                  
    x = "Season",
    y = "Number of Sales",
    breaks = 50,
    border = "black"
  ) +
theme_minimal()
```
From the graph we can see that the least amount of sales occur during the winter quarter, and the sales steadily increase throughout the year from spring, to summer, to the fall where the sales reach their highest volume. The least amount of sales occurs in the winter quarter and there are a number of reasons that may contribute to this. With colder and wetter weather potential buyers and sellers are less likely to want to move or sell their house during this period as it adds increased complexity to the viewing and moving period as there are less hours in a day and a higher likelihood of weather disturbance. This leads to a decrease in both the potential sellers and buyers on the market. Additionally, Christmas and New Years are also during the winter months when people tend to prefer to stay home and celebrate, rather than be moving or unpacking during the festive season.
During the period of increase from spring to the summer, the weather begins to improve meaning the days are getting longer and the weather is getting warmer, making viewing and moving much easier. Generally, people take time off during this period already making moving easier. Schools are also on break during the summer, which allows for further ease in viewing and relocating. The peak of the sales in the fall has the weather factors of the spring and summer, but is also prime time for relocation, as it is often when rent leases end, job cycles begin and schools begin to start up. This means that there is an increase in the demand for people looking to buy and sell houses.
```{r}
7.
yearly_summary <- bayside_residential_greater_than_0 %>%
  filter(!is.na(SALE_DATE)) %>%
  mutate(SALE_PRICE = as.numeric(gsub("[^0-9.]", "", SALE_PRICE))) %>%
  group_by(SALE_YEAR = as.numeric(format(SALE_DATE, "%Y"))) %>%
  summarise(
    total_sales = n(),
    total_revenue = sum(SALE_PRICE, na.rm = TRUE)
  )

earliest_date <- min(bayside_residential_greater_than_0$SALE_DATE, na.rm = TRUE)
latest_date <- max(bayside_residential_greater_than_0$SALE_DATE, na.rm = TRUE)

ggplot(yearly_summary, aes(x = SALE_YEAR, y = total_revenue)) +
  geom_line(color = "green", linewidth = 1) +
  geom_point() +
  labs(
    title = "Total Residential Revenue in Bayside by Year",
    subtitle = paste(
      format(earliest_date, "%b %d, %Y"), "to", format(latest_date, "%b %d, %Y")
    ),
    x = "Year",
    y = "Total Revenue ($)"
  ) +
  scale_y_continuous(labels = dollar) +
  theme_minimal()
```
From 2005-2007, we can see a moderate increase in total revenue for that period of time, from around $400 million to $550 million. In 2008 there was a sudden decrease in total revenue, which continued to decline until 2011, down to below $300 million.  From 2011 to 2017, the total revenue increased at a strong rate to above $600 million. From 2017 through to 2020, there is a moderate decrease to $440 million. In 2021 we can see the peak of the total revenues of this data, where it reached $800 million. From 2021 to 2022 there is a small decrease, before dropping drastically to below $500 million. Generally, across the 20 years, there has been a slight increase, but the fluctuations of decreasing and increasing periods are most notable. These are linked to the macroeconomic environment, with the most notable events being the 2008 global financial crisis and the 2020 covid pandemic, and their resulting impact on interest rates, inflation and disposable income. 

The opportunities and risks that are apparent come from the fluctuations that occur throughout this period. From an opportunity perspective, the periods of growth in revenue where economic sentiment is high means that it is a good time to be selling houses and generating revenue and growing profits. However, in the periods of negative growth where the economic sentiment is low and there are notable macroeconomic  and global events, there is increased risk as sales revenue drastically increases, making profits harder to attain. The property sales market is sensitive to economic factors, providing both opportunities and risk.